---
title: "Post title"
description: "One or two sentences that appear on the Applications listing card."
author: "Your name"
date: 2026-09-16
categories: [PCA, biplot]
draft: true
---

## The data

Introduce the data set. Where does it come from, what does one row represent, and which variables are we looking at?

## The question

State in plain language what we want to find out.

## The biplot

```{r}
#| label: setup
library(biplotEZ)
```

```{r}
#| label: biplot
#| fig-cap: "Caption explaining what the reader is looking at."
# biplot(data) |> PCA() |> plot()
```

## What the picture tells us

Walk the reader through the plot. Which observations cluster? Which variables pull them apart? What is surprising?

## Take-aways

- One-line summary of the first finding.
- One-line summary of the second finding.

::: {.callout-note collapse="true"}
## Reproduce this post
The full source of this page lives in the site repository. Install biplotEZ with `install.packages("biplotEZ")` and run the code chunks above.
:::
